\bf Designing Modular Software: A Case Study in Introductory Statistics

### 2016-10-20 jcgs-paper-2016/paper.tex
We have submitted our revision!

\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
    \usepackage{xltxtra,xunicode}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage{color}
\usepackage{fancyvrb}
\newcommand{\VerbBar}{|}
\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
% Add ',fontsize=\small' for more characters per line
\usepackage{framed}
\definecolor{shadecolor}{RGB}{248,248,248}
\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{{#1}}}}
\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{{#1}}}
\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{{#1}}}
\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{{#1}}}
\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{{#1}}}
\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{{#1}}}
\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{{#1}}}
\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{{#1}}}
\newcommand{\StringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{{#1}}}
\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{{#1}}}
\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{{#1}}}
\newcommand{\ImportTok}[1]{{#1}}
\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{{#1}}}}
\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{{#1}}}}}
\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{{#1}}}}}
\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{{#1}}}}}
\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{{#1}}}
\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{{#1}}}
\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{{#1}}}
\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{{#1}}}}
\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{{#1}}}}
\newcommand{\BuiltInTok}[1]{{#1}}
\newcommand{\ExtensionTok}[1]{{#1}}
\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{{#1}}}}
\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.77,0.63,0.00}{{#1}}}
\newcommand{\RegionMarkerTok}[1]{{#1}}
\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{{#1}}}}}
\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{{#1}}}}}
\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{{#1}}}
\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.64,0.00,0.00}{\textbf{{#1}}}}
\newcommand{\NormalTok}[1]{{#1}}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={Introductory statistics with intRo},
            colorlinks=true,
            citecolor=blue,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\setcounter{secnumdepth}{5}

@@

\title{\bf Introductory statistics with \texttt{intRo}}
\author{Eric Hare \\ Iowa State University \\ \texttt{}  and \\ Andee Kaplan \\ Iowa State University \\ \texttt{} }
\date{}

@@

\begin{abstract}
\texttt{intRo} is a web-based application for performing basic data
analysis and statistical routines. Leveraging the power of R and Shiny,
\texttt{intRo} implements common statistical functions in an extensible
modular structure, while including a point-and-click interface for the
novice statistician. This simplicity lends itself to a natural
presentation in an introductory statistics course as a substitute for
other commonly used statistical software packages. \texttt{intRo} is
currently deployed at the URL \url{http://www.intro-stats.com}. Within
this paper, we introduce the application and explore the design
decisions underlying \texttt{intRo}, as well as highlight some
challenges and advantages of reactive programming.
\end{abstract}

\noindent%
{\it Keywords:}  Interactivity, Programming Paradigms, Reactive Programming,
Reproducibility, Statistical Software
\vfill

@@

\section{Introduction}\label{introduction}

The widespread adoption of R (R Core Team 2014) as a tool for
statistical analysis has undoubtedly been an important development for
the scientific community. However, using R in most cases still requires
a basic knowledge of programming concepts which may pose a steep
learning curve for the introductory statistics student (Tan, Ting, and
Ling 2009). This additional time commitment may explain why introductory
courses often utilize point-and-click applications, even if the
instructor himself/herself uses R in their own work. Still, some
compromises must be made when using many graphical applications,
including dealing with software licenses and unsupported desktop
platforms.

@@

Multiple software packages have recently been written in an attempt to
spur interest in R programming and statistics. DataCamp's (DataCamp
2014) courses are user-friendly ways to learning basic R programming and
data analysis techniques. Swirl (Carchedi et al. 2014) is a similar
interactive tool to make learning R more fun by learning it within R
itself. Project MOSAIC (Pruim, Kaplan, and Horton 2014) has created a
suite of tools to simplify the teaching of statistics in the form of an
R package. The primary goal of DataCamp and Swirl is to teach R
programming, rather than facilitate the learning of introductory
statistics. Project MOSAIC's goal is to faciliate this learning, but
using the package requires a knowledge of R programming that the
introductory student may not have. R Commander (Fox 2005) and Deducer
(Fellows 2012) provide a graphical front-end to many statistical
functions in R, but are not web-based and thus require local
installation and configuration. iNZight Lite (Wild 2015) also attempts
to expose students to data analysis without requiring programming
knowledge, but does not include reproducible R code, and therefore has
less of a focus on spurring interest in coding for students.

Upon the release of RStudio's Shiny (RStudio and Inc. 2014) it became
easier for an R-based analysis to be converted to an interactive web
application. This in turn led us to create an introductory statistics
application which we call \texttt{intRo}, available at
\url{http://www.intro-stats.com}. \texttt{intRo} offers a number of key
advantages over traditional statistics software, including ease of
access and an aim to foster student interest in coding. Attempting to
entirely hide the programming aspect from students, even in introductory
classes, is a lost opportunity to get students interested in statistical
computing. It is also a lost opportunity reaching students who learn
differently or have a computational background. Another advantage is its
modular structure, which allows course instructors to tailor the
application towards the needs of a particular class, rather than accept
a piece of software as is. Additionally, \texttt{intRo} stands apart
from new tools in that it is a supplement to an existing class, fully
usable by a beginning statistics student. These advantages will be
discussed at length in this text. The paper is structured as follows:
section \ref{what-is-intro} introduces the application, its features and
its usability, and provides motivations for the design decisions
underlying it. Section \ref{understanding-intro} provides technical
details on the design decisions of \texttt{intRo}, in particular, its
modular and extensible structure. Section
\ref{deploying-intro-instances} describes the process of deploying a
classroom instance of \texttt{intRo} tailored to the individual
instructor. Finally, section \ref{conclusions-and-future-work} discusses
some future possibilities and limitations of the software.

\section{\texorpdfstring{What is
\texttt{intRo}?}{What is intRo?}}\label{what-is-intro}

\texttt{intRo} is a web-based tool to accompany an introductory
statistics class. It is meant to assist in the learning of statistics
rather than as a stand-alone delivery of statistics education, with the
intention of being used in conjunction with a guided class. An
accompanying R package, titled \texttt{intRo} and available on GitHub,
assists in the downloading, running, and deploying of \texttt{intRo}
instances.

@@

In the development of \texttt{intRo}, we focused on aspects of the user
interface (UI) and output that make it easy to pick up without extensive
training. We used large, easy to click icons in the page header to help
students find what they need more easily. We also made the functionality
available the minimal necessary for an introductory statistics course.
Figure \ref{fig:user_experience} illustrates the simple steps a student
takes to generate a result in \texttt{intRo}. In this instance, a
student clicks on the Graphical tab to create a mosaic plot. The student
sees the plot, and elects to click the save button to store the plot and
its corresponding code to the final compendium.

\begin{figure}[H]
\centering
\includegraphics[width=\linewidth]{user_experience.pdf}
\caption{A typical student experience of generating a result in \texttt{intRo}. In this instance, a student clicks on the Graphical tab to create a mosaic plot. The student sees the plot, and elects to click the save button to store the plot (and its corresponding code) to the final compendium.}
\label{fig:user_experience}
\end{figure}

@@

\begin{figure*}[ht!]
\centering
\includegraphics[width=\linewidth]{ui_annotate.pdf}
\caption{The five elements that comprise the \texttt{intRo} application: 1) top navigation, 2) side navigation, 3) options panel, 4) results panel, and 5) code panel.}
\label{fig:ui}
\end{figure*}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  \textbf{Top Navigation} - The top navigation bar includes two sets of
  clickable icons. The left-aligned buttons are informational buttons.
  The first is a link to \texttt{intRo}. The second is a link to the
  documentation page. The third is a link to the GitHub repository where
  the code for \texttt{intRo} is housed. The final button is a link to
  our websites, which contain contact information if there are any
  questions or comments. The right-aligned buttons are \texttt{intRo}
  utilities. The first is a link to toggle the visibility of the code
  panel (5). The middle icon downloads an rmarkdown (Allaire et al.
  2014) document of the analysis performed. The last is a link to print
  the stored module results, and the associated code (if visible).
\item
  \textbf{Side Navigation} - The side navigation panel includes a list
  of data analysis tasks.
\item
  \textbf{Options Panel} - The options panel includes task-specific
  options which the student can use to customize their results.
\item
  \textbf{Results Panel} - The results pane displays the result of the
  selected module and options.
\item
  \textbf{Code Panel} - The code panel displays the R code used to
  generate the results from the student's \texttt{intRo} session. The
  code panel is shown by default to facilitate a transition to coding,
  but can be hidden by clicking the code toggle button in the Top
  Navigation bar.
\end{enumerate}

The modules included in \texttt{intRo} are split into three higher level
categories - data, summaries, and inference. Under each of these
categories, there are seven default modules, which perform specific data
analysis tasks that employ an easy to use point-and-click interface.
More modules can easily be added by an instructor, as detailed in
section \ref{modules}. The default modules support uploading and
downloading a dataset, transforming variables, graphical and numerical
summaries, simple linear regression, contingency tables, and T-tests.

We've also created a documentation website that is consistent with the
interface of the application. The documentation is available at
\url{http://gammarama.github.io/intRo} and hyperlinked within the
application itself. The documentation describes the specific
functionality provided by the default \texttt{intRo} modules.

Statistics courses with a lab component can most benefit from the use of
\texttt{intRo}. As students learn the aforementioned concepts, they can
practice using the tools to perform the statistical tests and summaries.
They simply upload a dataset and click the module appropriate to the
section of the class. Once they choose the options desired by the
instructor (variables, confidence levels, etc.), they can view the
results and ultimately print the output for turning in. Early labs can
walk the student through the limited number of button presses and clicks
necessary, while later labs can more vaguely describe the intended
output in hopes that the student can easily navigate the \texttt{intRo}
interface to obtain the results.

\texttt{intRo} has an ulterior motive as well: to get students excited
about programming. By navigating about the user interface of
\texttt{intRo}, students are actually creating a fully-executable R
script that they can download and run locally as well as viewing the
script change real-time within the application. This code creation
element of \texttt{intRo} is meant to generate excitement about
programming in R and empower students to feel that they can generate
code as well. \texttt{intRo} uses rmarkdown's render function in order
to print the results, by dymanically executing the student's R script.
By default, the output will include the R code, but if the student
elects to hide the source code by clicking the code toggle button at the
top, the code will not appear in the printed results.

This reproducibility framework has three key advantages. First, this
eases a student who may be intimidated by programming into the idea that
interacting with a user interface is really just a frontend for code.
Seeing the correspondence between graphical clicks and printed code will
hopefully also lessen the fear of coding that many students may have.
Second, an analysis created by \texttt{intRo} can be reproduced in an R
session to easily assess and extend the results. Finally, ``printing''
the results of an \texttt{intRo} analysis amounts to nothing more than
executing the R code on the server, adding another layer of
reproducibility. These concepts are important because they encourage
best practices with regards to disclosure of analysis methods in
research (Baggerly and Berry 2011; Xie 2015).

On the front end, user interaction with \texttt{intRo} is split into
bitesize chunks that we call modules. Modules are self-contained pieces
of functionality which implement common statistical procedures that are
often used in introductory statistics classes. These modules form the
core of \texttt{intRo} and are discussed at length in the next section.

\section{\texorpdfstring{Understanding
\texttt{intRo}}{Understanding intRo}}\label{understanding-intro}

To fully understand \texttt{intRo}, it is important to expand upon two
underlying concepts. The first is the idea of modularity in the context
of a Shiny application. The second is the idea of reactive programming
in the context of a web application.

\subsection{Modules}\label{modules}

An \texttt{intRo} module is a set of self-contained executible R scripts
that together produce a set of introductory statistics functionality.
\texttt{intRo} modules were designed in this way to allow for simple
dynamic creation of the user interface at run-time, as well as ease the
process of converting existing analysis code to the \texttt{intRo}
framework. A high-level diagram of this process is given in figure
\ref{fig:app_creation_modules}. \texttt{intRo} modules are split up into
multiple R scripts which are included either in Shiny's user interface
or server definitions. At runtime, the \texttt{intRo} sources in the
specified modules (contained in the modules folder) to dynamically
generate the functionality available in the application. This allows for
the specific functionality needed to be determined and adjusted by the
individual course instructor. In this example, the instructor is
electing to include a nonparametric module, which is not enabled by
default, to allow the students to perform a wilcoxon rank sum test.

>>>>

\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
    \usepackage{xltxtra,xunicode}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage{color}
\usepackage{fancyvrb}
\newcommand{\VerbBar}{|}
\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
% Add ',fontsize=\small' for more characters per line
\usepackage{framed}
\definecolor{shadecolor}{RGB}{248,248,248}
\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{{#1}}}}
\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{{#1}}}
\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{{#1}}}
\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{{#1}}}
\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{{#1}}}
\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{{#1}}}
\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{{#1}}}
\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{{#1}}}
\newcommand{\StringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{{#1}}}
\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{{#1}}}
\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{{#1}}}
\newcommand{\ImportTok}[1]{{#1}}
\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{{#1}}}}
\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{{#1}}}}}
\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{{#1}}}}}
\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{{#1}}}}}
\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{{#1}}}
\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{{#1}}}
\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{{#1}}}
\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{{#1}}}}
\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{{#1}}}}
\newcommand{\BuiltInTok}[1]{{#1}}
\newcommand{\ExtensionTok}[1]{{#1}}
\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{{#1}}}}
\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.77,0.63,0.00}{{#1}}}
\newcommand{\RegionMarkerTok}[1]{{#1}}
\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{{#1}}}}}
\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{{#1}}}}}
\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{{#1}}}
\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.64,0.00,0.00}{\textbf{{#1}}}}
\newcommand{\NormalTok}[1]{{#1}}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={Designing Modular Software: A Case Study in Introductory Statistics},
            colorlinks=true,
            citecolor=blue,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\setcounter{secnumdepth}{5}

@@

\title{\bf Designing Modular Software: A Case Study in Introductory Statistics}
\author{Eric Hare \\ Iowa State University \\ \texttt{}  and \\ Andee Kaplan \\ Iowa State University \\ \texttt{} }
\date{}

@@

\begin{abstract}
Modular programming is a development paradigm that emphasizes
self-contained, flexible, and independent pieces of functionality. This
practice allows new features to be seamlessly added when desired, and
unwanted features to be removed, thus simplifying the user-facing view
of the software. The recent rise of web-based software applications has
presented new challenges for designing an extensible, modular software
system. In this paper, we outline a framework for designing such a
system, with a focus on reproducibility of the results. We present as a
case study a Shiny-based web application called \texttt{intRo}, that
allows the user to perform basic data analyses and statistical routines.
Finally, we highlight some challenges we encountered, and how to address
them, when combining modular programming concepts with reactive
programming as used by Shiny.
\end{abstract}

\noindent%
{\it Keywords:}  Interactivity, Modularity, Programming Paradigms, Reactive Programming,
Reproducibility, Statistical Software
\vfill

@@

\section{Background}\label{background}

Modularity is a pervasive concept in computer science, extending from
the design of systems (Parnas 1972), to the design of software
(Szyperski 1996). Modularity offers several advantages to both a
developer and a user. In particular, functionality can be dynamically
loaded and unloaded depending on the particular use case. Open source
modular software precipitates the possibility of extensions contributed
by a wide array of programmers, which can allow the software to morph
into areas that weren't anticipated early in development. In the
statistics realm, R (R Core Team 2014) is a prime example of the virtues
of modular programming. As of this writing, The Comprehensive R Archive
Network (CRAN) contains over 9000 source packages which can be installed
and dynamically loaded in a particular session as needed.

Other statistics software also makes use of a number of these ideas.
Microsoft Excel and JMP both include support for extensions, called
macros and add-ins respectively, which allow programming routines to be
written extending the base functionality of these programs. Compared
with R, however, these programs don't maintain a large central
repository of publicly available extensions on the level of CRAN. There
are also software packages building upon R, and thus gaining the
advantages of CRAN natively, such as R Commander (Fox 2005) and Deducer
(Fellows 2012), which each provide a graphical front-end to many
statistical functions in R. One thing these software packages all have
in common is the requirement of local installation and configuration,
which means certain operating systems and platforms will not support
their use.

With the advantages of R clear, an approach to building statistical
software and statistical learning tools would be to attempt to generate
interest in programming, which could help naturally ease the transition
into the use of R. Multiple software packages have recently been written
in an attempt to spur this interest in R programming and statistics.
DataCamp's (DataCamp 2014) courses are a user-friendly way to learn
basic R programming and data analysis techniques. Swirl (Carchedi et al.
2014) is a similar interactive tool to make learning R more fun by
learning it within R itself. Project MOSAIC (Pruim, Kaplan, and Horton
2014) has created a suite of tools to simplify the teaching of
statistics in the form of an R package. The primary goal of DataCamp and
Swirl is to teach R programming, rather than facilitate the learning of
introductory statistics.

Modern web technologies have enabled a new generation of software
packages that reside solely on the web, which eliminates the issue of
local installation and helps abstract away some of the more challenging
programming aspects of working directly with R. Upon the release of
RStudio's Shiny (RStudio and Inc. 2014) it became easier for an R-based
analysis to be converted to an interactive web application. Several
recent software packages have built upon Shiny to provide a web-based
system based on R. One such package is iNZight Lite (Wild 2015) which
attempts to expose students to data analysis without requiring
programming knowledge. Like most web-based systems, this does not
include reproducible R code, which limits its usefulness in a scientific
or academic setting. Another package is called Radiant (Nijs 2016),
which is a web-based application with the aim of furthering business
education and financial analysis. While the application is modular and
extensible, it does require installation and hosting and is inundated
with more features than necessary for an introductory student. An
overview of the comparison between the features of these statistical
software packages is presented in Table \ref{tab:compare}. Partial
fulfillment of requirements is noted in the table, as well as a measure
of the complexity of functionality offered by default. For example,
\texttt{R} does have an associated Graphical User Interface (GUI),
however this interface is very limited, thus only partially fulfilling
the behavior of a GUI.

\begin{table}[ht]
\centering
\begin{tabular}{llllllll}
  \hline
Software & GUI & Install & Modular & Web & Extensible & Reproducible & Features \\ 
  \hline
intRo & Yes & No & Yes & Yes & Yes & Yes & Limited \\ 
  JMP & Yes & Yes & Partial & No & Partial & No & Full \\ 
  R & Partial & Yes & Yes & No & Yes & Yes & Full \\ 
  Rcmdr & Yes & Yes & No & No & Yes & Yes & Moderate \\ 
  Deducer & Yes & Yes & No & No & Yes & Yes & Moderate \\ 
  MOSAIC & No & Yes & No & No & No & Yes & Limited \\ 
  iNZight Lite & Yes & No & Yes & Yes & Partial & No & Limited \\ 
  Radiant & Yes & Yes & Yes & Yes & Partial & Yes & Moderate \\ 
   \hline
\end{tabular}
\caption{A comparison of statistical software packages across the metrics of usability, modularity, extensibility, and reproducibility. Partial fulfillment of requirements is noted in the table, as well as a measure of the complexity of functionality offered by default.} 
\label{tab:compare}
\end{table}

Though challenging in a GUI, a reproducibility framework has three key
advantages. First, it eases a student who may be intimidated by
programming into the idea that interacting with a user interface is
really just a frontend for code. Seeing the correspondence between
graphical clicks and printed code should help lessen the fear of coding
that many students may have. Second, an analysis created by a
reproducible software system can be brought in an R session to easily
assess and extend the results. Finally, with the help of knitr (Xie
2015) and rmarkdown (Allaire et al. 2014), ``printing'' the results of a
reproducible software system analysis amounts to nothing more than
executing the R code on the server, adding another layer of
reproducibility. These concepts are important because they encourage
best practices with regards to disclosure of analysis methods in
research (Baggerly and Berry 2011; Xie 2015).

Based on the above, we believe a modern software system should be
\textbf{modular}, \textbf{extensible}, \textbf{web-based}, and foster
\textbf{reproducibility}. We have developed a case-study application
called \texttt{intRo} which we will use to illustrate our method of
developing a system meeting these criteria. The paper is structured as
follows: Section \ref{case-study-intro} introduces the application, its
features and its usability, and provides motivations for why it was
built. Section \ref{intro-design-decisions} provides technical details
on how we built \texttt{intRo}, by walking through the underlying
modularity, reproducibility, and reactive framework, as well as how it
can be used to develop other software systems with these properties.
Finally, Section \ref{conclusions-and-future-work} discusses some future
possibilities and limitations of both \texttt{intRo}, and modular
systems in general.

\section{\texorpdfstring{Case Study:
\texttt{intRo}}{Case Study: intRo}}\label{case-study-intro}

The widespread adoption of R as a tool for statistical analysis has
undoubtedly been an important development for the scientific community.
However, using R in most cases still requires a basic knowledge of
programming concepts which may pose a steep learning curve for the
introductory statistics student (Tan, Ting, and Ling 2009). This
additional time commitment may explain why introductory courses often
utilize point-and-click applications, even if the instructor
himself/herself uses R in their own work. Still, some compromises must
be made when using many graphical applications, including dealing with
software licenses and unsupported desktop platforms. From the
instructor's perspective in particular, managing a large group of
software licenses for students with various computing environments and
versions could wind up being extremely cumbersome.

@@

These challenges led us to create an introductory statistics application
which we call \texttt{intRo}, available at
\url{http://www.intro-stats.com}. \texttt{intRo} offers a number of key
advantages over traditional statistics software, including ease of
access and an aim to foster student interest in coding. Attempting to
entirely hide the programming aspect from students, even in introductory
classes, is a lost opportunity to get students interested in statistical
computing. It is also a lost opportunity reaching students who learn
differently or have a computational background. Another advantage is its
modular structure, which allows course instructors to tailor the
application towards the needs of a particular class, rather than accept
a piece of software as is. Additionally, \texttt{intRo} stands apart
from new tools in that it is a supplement to an existing class, fully
usable by a beginning statistics student. An accompanying R package,
titled \texttt{intRo} and available on GitHub, assists in the
downloading, running, and deploying of \texttt{intRo} instances (See
Section \ref{deploying-intro-instances}).

@@

In the development of \texttt{intRo}, we focused on aspects of the user
interface (UI) and output that make it easy to pick up without extensive
training. We used large, easy to click icons in the page header to help
students find what they need more easily. We also made the functionality
available the minimal necessary for an introductory statistics course.
Figure \ref{fig:user_experience} presents a schematic of the simple
steps a student takes to generate a result in \texttt{intRo}. In this
instance, a student clicks on the Graphical tab to create a mosaic plot.
The student sees the plot, and elects to click the save button to store
the plot and its corresponding code to the final compendium.

\begin{figure}[H]
\centering
\includegraphics[width=\linewidth]{images/user_experience.pdf}
\caption{A schematic of the typical student experience of generating a result in \texttt{intRo}. In this instance, a student clicks on the Graphical tab to create a mosaic plot. The student sees the plot, and elects to click the save button to store the plot (and its corresponding code) to the final compendium.}
\label{fig:user_experience}
\end{figure}

@@

\begin{figure*}[ht!]
\centering
\includegraphics[width=\linewidth]{images/ui_annotate.pdf}
\caption{The five elements that comprise the \texttt{intRo} application: 1) top navigation, 2) side navigation, 3) options panel, 4) results panel, and 5) code panel.}
\label{fig:ui}
\end{figure*}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  \textbf{Top Navigation} - The top navigation bar includes two sets of
  clickable icons. The left-aligned buttons are informational buttons.
  The first is a link to \texttt{intRo}. The second is a link to the
  documentation page. The third is a link to the GitHub repository where
  the code for \texttt{intRo} is housed. The final button is a link to
  our websites, which contain contact information if there are any
  questions or comments. The right-aligned buttons are \texttt{intRo}
  utilities. The first is a link to toggle the visibility of the code
  panel (5). The middle icon downloads an rmarkdown document of the
  analysis performed. The last is a link to print the stored module
  results, and the associated code (if visible).
\item
  \textbf{Side Navigation} - The side navigation panel includes a list
  of data analysis tasks.
\item
  \textbf{Options Panel} - The options panel includes task-specific
  options which the student can use to customize their results.
\item
  \textbf{Results Panel} - The results pane displays the result of the
  selected module and options.
\item
  \textbf{Code Panel} - The code panel displays the R code used to
  generate the results from the student's \texttt{intRo} session. The
  code panel is shown by default to facilitate a transition to coding,
  but can be hidden by clicking the code toggle button in the Top
  Navigation bar.
\end{enumerate}

The modules included in \texttt{intRo} are split into three higher level
categories - data, summaries, and inference. Under each of these
categories, there are seven default modules, which perform specific data
analysis tasks that employ an easy to use point-and-click interface.
More modules can easily be added by an instructor, as detailed in
Section \ref{designing-for-modularity}. The default modules support
uploading and downloading a dataset, transforming variables, graphical
and numerical summaries, simple linear regression, contingency tables,
and T-tests.

\texttt{intRo} has an ulterior motive as well: to get students excited
about programming. By navigating about the user interface of
\texttt{intRo}, students are actually creating a fully-executable,
reproducible R script that they can download and run locally as well as
viewing the script change real-time within the application. This code
creation element of \texttt{intRo} is meant to generate excitement about
programming in R and empower students to feel that they can generate
code as well. \texttt{intRo} uses rmarkdown's render function in order
to print the results, by dymanically executing the student's R script.
By default, the output will include the R code, but if the student
elects to hide the source code by clicking the code toggle button at the
top, the code will not appear in the printed results.

On the front end, user interaction with \texttt{intRo} is split into
bitesize chunks that we call modules. In \texttt{intRo}'s context,
modules are self-contained pieces of functionality which implement
common statistical procedures. These modules form the core functionality
of \texttt{intRo} and are discussed at length in the next section.

\section{\texorpdfstring{\texttt{intRo} Design
Decisions}{intRo Design Decisions}}\label{intro-design-decisions}

In this section, we detail the design choices surrounding
\texttt{intRo}'s extensibility. We have designed it in such a way that
these ideas can be used in other Shiny-based software projects.

\subsection{Designing for Modularity}\label{designing-for-modularity}

An \texttt{intRo} module is a set of self-contained executible R scripts
that together produce a set of introductory statistics functionality.
\texttt{intRo} modules were designed in this way to allow for simple
dynamic creation of the user interface at run-time, as well as ease the
process of converting existing analysis code to the \texttt{intRo}
framework. A high-level diagram of this process is given in Figure
\ref{fig:app_creation_modules}. \texttt{intRo} modules are split up into
multiple R scripts which are included either in Shiny's user interface
or server definitions. At runtime, the \texttt{intRo} sources in the
specified modules (contained in the modules folder) to dynamically
generate the functionality available in the application. This allows for
the specific functionality needed to be determined and adjusted by the
individual course instructor. In this example, the instructor is
electing to include a nonparametric module, which is not enabled by
default, to allow the students to perform a wilcoxon rank sum test.

\begin{figure*}[ht!]
\centering
\includegraphics[width=\linewidth]{images/app_creation_modules.pdf}
\caption{This figure depicts how the Shiny \texttt{server.R} and \texttt{ui.R} files are populated using the modular structure within \texttt{intRo}. \texttt{intRo} modules are split up into multiple R scripts which are included either in Shiny's user interface or server definitions. At runtime, the \texttt{intRo} sources in the specified modules (contained in the modules folder) to dynamically generate the functionality available in the application. This allows for the specific functionality needed to be determined and adjusted by the individual course instructor. In this example, the instructor is electing to include a nonparametric module, which is not enabled by default, to allow the students to perform a wilcoxon rank sum test.}
\label{fig:app_creation_modules}
\end{figure*}
